---
subtitle: "Data Science 306: Lecture 28"
title: "Databases and SQL from R"
output:
  learnr::tutorial:
    progressive: true
    css: css/lecture.css
runtime: shiny_prerendered
---


```{r setup, include=FALSE}
library(learnr)
library(tidyverse)
library(DBI)
library(RSQLite)        # install.packages("RSQLite")
library(duckdb)         # install.packages("duckdb")
library(dbplyr)         # install.packages("dbplyr")
library(nycflights13)   # install.packages("nycflights13")
theme_set(theme_bw())

# `con` lives here (not in a slide) because learnr exercises inherit only the
# setup chunk's environment, and the exercises run against `con`. The DuckDB
# connection is created later in its own slide.
con <- dbConnect(RSQLite::SQLite(), ":memory:")
dbWriteTable(con, "mpg",              ggplot2::mpg)
dbWriteTable(con, "band_members",     dplyr::band_members)
dbWriteTable(con, "band_instruments", dplyr::band_instruments)
dbWriteTable(con, "flights",          nycflights13::flights)
dbWriteTable(con, "airlines",         nycflights13::airlines)
```

## Overview

Our data has lived in files read entirely into memory. When that stops working:

* **Databases** -- where most real data lives, and why
* **SQL** -- the standard query language, across databases *and* languages
* **`DBI`** -- connect from R and send SQL
* **`dbplyr`** -- `dplyr` that runs *inside* the database
* **Two engines** -- **SQLite**, then **DuckDB** (queries **Parquet** directly);
  plus the same SQL run from **Python**

## Why a database?

A **database** is a separate program built for data:

* **Size** -- bigger than RAM; reads only what a query needs
* **Efficiency** -- structured for fast filters and joins
* **Persistence** -- lives on disk, not your R session
* **Concurrency** -- many users read/write at once
* **One source of truth** -- no more `final_data_v3_REAL.csv` by email
* **Distributed** -- can be split across machines (*sharding*)
* **Standardized** -- many languages share the same data

Most are **relational**: tables linked by keys -- our *joins*.

## Talking to a database: `DBI`

R talks to any database through **`DBI`** plus a **backend** package:

* `RSQLite` -- tiny, self-contained, in a single file; what we use today
* `RPostgres`, `RMariaDB`, `bigrquery` -- production databases and warehouses

Same interface regardless of backend: connect, work, disconnect.

```{r}
# open a brand-new, empty in-memory SQLite database
tmp_con <- dbConnect(RSQLite::SQLite(), ":memory:")

# copy a tibble into it as a table
dbWriteTable(tmp_con, "mpg", ggplot2::mpg)

# what tables exist now?
dbListTables(tmp_con)

dbDisconnect(tmp_con)   # always close a connection when finished
```

`con` (used all lecture) was opened the same way behind the scenes. It holds:

```{r}
dbListTables(con)
```

## Sending SQL directly

**SQL** is the database's native language; send it with `dbGetQuery()`:

```{r}
dbGetQuery(con, "
  SELECT manufacturer, AVG(hwy) AS mean_hwy
  FROM mpg
  WHERE year = 2008
  GROUP BY manufacturer
  ORDER BY mean_hwy DESC
")
```

Same shape as a `dplyr` pipeline.

## Anatomy of a `SELECT`

One `SELECT` statement, clauses in a fixed order:

```sql
SELECT   columns and summaries you want back
FROM     which table
WHERE    keep only rows matching a condition
GROUP BY split rows into groups
ORDER BY sort the result
LIMIT    keep only the first n rows
```

Only `SELECT` and `FROM` are required:

```{r}
dbGetQuery(con, "SELECT manufacturer, model, hwy FROM mpg LIMIT 5")
```

`SELECT *` -- every column:

```{r}
dbGetQuery(con, "SELECT * FROM mpg LIMIT 3")
```

## `WHERE`: keeping rows

`WHERE` is `filter()` -- keep matching rows (`=`, not `==`; combine with
`AND` / `OR`):

```{r}
dbGetQuery(con, "
  SELECT manufacturer, model, year, hwy
  FROM mpg
  WHERE hwy > 30 AND year = 2008
")
```

Handy idioms: `IN (...)` (like `%in%`), `BETWEEN a AND b`, `LIKE` (`%` = any
characters):

```{r}
dbGetQuery(con, "
  SELECT DISTINCT manufacturer
  FROM mpg
  WHERE manufacturer IN ('audi', 'honda')
     OR manufacturer LIKE 'to%'
")
```

## Exercise: `SELECT` + `WHERE`

We'll grow one query over the next few exercises. To start: from `mpg`, select
`manufacturer`, `model`, and `hwy`, keeping only **2008 cars with `hwy` of at
least 30**.

```{r ex-where, exercise = TRUE}
dbGetQuery(con, "
  SELECT...") 

```{r ex-where-solution}
dbGetQuery(con, "
  SELECT manufacturer, model, hwy
  FROM mpg
  WHERE year = 2008 AND hwy >= 30
")
```

## `GROUP BY` and aggregates

`GROUP BY` + **aggregates** (`COUNT(*)`, `SUM`, `AVG`, `MIN`, `MAX`) is
`group_by() |> summarize()`:

```{r}
dbGetQuery(con, "
  SELECT class,
         COUNT(*)  AS n,
         AVG(hwy)  AS mean_hwy,
         MIN(hwy)  AS worst,
         MAX(hwy)  AS best
  FROM mpg
  GROUP BY class
  ORDER BY mean_hwy DESC
")
```

## Exercise: group and summarize

Now summarize instead of listing rows: for each `class`, return the count and
the (rounded) average `hwy`. Hint: remember `COUNT`, `ROUND`, and `AVG`.

```{r ex-groupby, exercise = TRUE}
dbGetQuery(con, "

")
```

```{r ex-groupby-solution}
dbGetQuery(con, "
  SELECT class, COUNT(*) AS n, ROUND(AVG(hwy), 1) AS mean_hwy
  FROM mpg
  GROUP BY class
")
```

## `WHERE` vs `HAVING`

`WHERE` filters rows *before* grouping; `HAVING` filters groups *after*:

```{r}
dbGetQuery(con, "
  SELECT class, COUNT(*) AS n, AVG(hwy) AS mean_hwy
  FROM mpg
  GROUP BY class
  HAVING n > 30
")
```

`HAVING` = a `filter()` after `summarize()`.

## Exercise: filter the groups

Extend your class summary: keep only classes with **more than 30 cars**, sorted
by average `hwy` (highest first).

```{r ex-having, exercise = TRUE}
dbGetQuery(con, "
  SELECT class, COUNT(*) AS n, ROUND(AVG(hwy), 1) AS mean_hwy
  FROM mpg
  GROUP BY class
  -- keep only big classes, then sort
")
```

```{r ex-having-solution}
dbGetQuery(con, "
  SELECT class, COUNT(*) AS n, ROUND(AVG(hwy), 1) AS mean_hwy
  FROM mpg
  GROUP BY class
  HAVING n > 30
  ORDER BY mean_hwy DESC
")
```

## Building up: common table expressions (`WITH`)

`WITH name AS (...)` names an intermediate result instead of nesting -- reads
top-to-bottom, like piped steps:

```{r}
dbGetQuery(con, "
  WITH class_stats AS (
    SELECT class, COUNT(*) AS n, AVG(hwy) AS mean_hwy
    FROM mpg
    GROUP BY class
  )
  SELECT class, mean_hwy
  FROM class_stats
  WHERE n > 30
  ORDER BY mean_hwy DESC
")
```

## Ranking within groups: window functions

A **window function** computes across a group but keeps **every row**
(`OVER (PARTITION BY ...)`) -- e.g. rank within group:

```{r}
dbGetQuery(con, "
  SELECT manufacturer, model, hwy,
         RANK() OVER (PARTITION BY manufacturer ORDER BY hwy DESC) AS hwy_rank
  FROM mpg
  WHERE manufacturer IN ('toyota', 'honda')
  ORDER BY manufacturer, hwy_rank
  LIMIT 8
")
```

Like `group_by() |> mutate()` -- the rows survive (unlike `summarize()`).

## Exercise: rank within a group

Rank cars by `hwy` **within each `class`** (highest = rank 1). Return `class`,
`model`, `hwy`, and the rank.

```{r ex-window, exercise = TRUE}
dbGetQuery(con, "

")
```

```{r ex-window-solution}
dbGetQuery(con, "
  SELECT class, model, hwy,
         RANK() OVER (PARTITION BY class ORDER BY hwy DESC) AS hwy_rank
  FROM mpg
  ORDER BY class, hwy_rank
")
```

## `SELECT` is `dplyr` in a different order

| `dplyr`                  | SQL                                          |
|--------------------------|----------------------------------------------|
| `select()`               | `SELECT`                                     |
| `filter()`               | `WHERE`                                      |
| `group_by()`             | `GROUP BY`                                   |
| `summarize()`            | aggregates in `SELECT` (`AVG`, `COUNT`, ...) |
| `filter()` after summarize | `HAVING`                                   |
| `arrange()`              | `ORDER BY`                                   |
| `arrange(desc(x))`       | `ORDER BY x DESC`                            |
| `head(n)` / `slice()`    | `LIMIT n`                                    |
| `mutate()` / `rename()`  | expressions / `AS` in `SELECT`               |
| `left_join()` etc.       | `... JOIN ... ON ...`                         |
  
* **Declarative** -- you describe the result; the *query planner* picks how.
* Written `SELECT…FROM…WHERE…GROUP BY…HAVING…ORDER BY…LIMIT`, but *run*
  `FROM → WHERE → GROUP BY → HAVING → SELECT → ORDER BY → LIMIT`.

## Why filters and joins are fast: indexes

An **index** lets the database jump to matching rows instead of scanning.
`EXPLAIN QUERY PLAN` shows the plan -- no index means a full **scan**:

```{r}
dbGetQuery(con, "EXPLAIN QUERY PLAN SELECT * FROM flights WHERE origin = 'EWR'")
```

Add an index and ask again:

```{r}
dbExecute(con, "CREATE INDEX idx_origin ON flights(origin)")
dbGetQuery(con, "EXPLAIN QUERY PLAN SELECT * FROM flights WHERE origin = 'EWR'")
```

**`SCAN`** → **`SEARCH ... USING INDEX`**: same answer, no full scan. (Costs
storage and upkeep.)

## Joining tables in SQL

Data split across tables linked by keys; `JOIN ... ON` recombines them:

```{r}
dbGetQuery(con, "SELECT * FROM band_members")
dbGetQuery(con, "SELECT * FROM band_instruments")
```

**Inner join** -- keep only matches; `ON` names the key:

```{r}
dbGetQuery(con, "
  SELECT band_members.name, band_members.band, band_instruments.plays
  FROM band_members
  JOIN band_instruments
    ON band_members.name = band_instruments.name
")
```

= `inner_join(band_members, band_instruments, by = \"name\")`.

## Left joins and table aliases

`LEFT JOIN` keeps every left row (`NA` where no match); `AS` gives a table a
short alias:

```{r}
dbGetQuery(con, "
  SELECT m.name, m.band, i.plays
  FROM band_members AS m
  LEFT JOIN band_instruments AS i
    ON m.name = i.name
")
```

Join key → `ON`; `JOIN` / `LEFT JOIN` = `inner_join()` / `left_join()`.

## A realistic join

Same pattern at scale: `flights.carrier` → `airlines.name`, then aggregate:

```{r}
dbGetQuery(con, "
  SELECT a.name, COUNT(*) AS n_flights, AVG(f.dep_delay) AS mean_delay
  FROM flights AS f
  JOIN airlines AS a
    ON f.carrier = a.carrier
  GROUP BY a.name
  ORDER BY mean_delay DESC
  LIMIT 5
")
```

Scanned all 336k rows, returned 5.

## Exercise: join two tables

Inner-join `flights` to `airlines` on `carrier`. For each airline `name`, return
the flight count and average `arr_delay`, **best (lowest) average first**, top 5.

```{r ex-join, exercise = TRUE}
dbGetQuery(con, "
  SELECT a.name, ___
  FROM flights AS f
  JOIN airlines AS a
    ON ___
  GROUP BY a.name
  ___
")
```

```{r ex-join-solution}
dbGetQuery(con, "
  SELECT a.name, COUNT(*) AS n, AVG(f.arr_delay) AS mean_arr_delay
  FROM flights AS f
  JOIN airlines AS a
    ON f.carrier = a.carrier
  GROUP BY a.name
  ORDER BY mean_arr_delay ASC
  LIMIT 5
")
```

We could keep writing SQL by hand -- or let `dplyr` write it.

## `dbplyr`: `dplyr` that runs in the database

Point `tbl()` at a table, write `dplyr`; it's **translated to SQL** and run in
the database:

```{r}
# a reference to the table -- NOT a copy of the data
mpg_db <- tbl(con, "mpg")

result <- mpg_db |>
  filter(year == 2008) |>
  group_by(manufacturer) |>
  summarize(mean_hwy = mean(hwy, na.rm = TRUE)) |>
  arrange(desc(mean_hwy))
```

Same code as on a tibble -- but notice: **no output** (next slide).

## Lazy evaluation: nothing happens until you ask

`dbplyr` is **lazy** -- building a pipeline runs nothing. It executes when you:

* **print** it -- runs the query, previews a few rows
* **`collect()`** -- pulls the *full* result into R as a tibble

```{r}
result            # builds + previews a few rows, still "in the database"
```

```{r}
local_tbl <- result |> collect()   # NOW bring the whole result into R
class(local_tbl)                   # an ordinary tibble we can plot, model, ...
local_tbl
```

Laziness lets `dbplyr` fold the whole pipeline into **one** SQL query.

## Seeing the generated SQL

`show_query()` prints the SQL it will send -- great for learning and checking:

```{r}
result |> show_query()
```

Joins translate too:

```{r}
tbl(con, "flights") |>
  left_join(tbl(con, "airlines"), by = "carrier") |>
  group_by(name) |>
  summarize(mean_delay = mean(dep_delay, na.rm = TRUE)) |>
  show_query()
```

## Push work to the database, pull small results back

**Do the work in the database; `collect()` only the small result.**

```{r}
# GOOD: shrink in the database, then collect ~15 rows
summary_tbl <- tbl(con, "flights") |>
  filter(dest == "DTW") |>
  group_by(origin) |>
  summarize(n = n(), mean_delay = mean(arr_delay, na.rm = TRUE)) |>
  collect()

summary_tbl
```

```{r}
# BAD: collect all 336k rows first, then shrink (may not even fit)
bad <- tbl(con, "flights") |>
  collect() |>
  filter(dest == "DTW") |>
  group_by(origin) |>
  summarize(n = n(), mean_delay = mean(arr_delay, na.rm = TRUE))
```

Same answer -- but `collect()` belongs at the **end**, after the data is small.

## Exercise: fix the slow pipeline

This `collect()`s the whole table first. Reorder it so the database filters and
summarizes, with `collect()` **last**.

```{r ex-reorder, exercise = TRUE}
tbl(con, "flights") |>
  collect() |>
  filter(dest == "DTW") |>
  summarize(mean_delay = mean(arr_delay, na.rm = TRUE))
```

```{r ex-reorder-solution}
tbl(con, "flights") |>
  filter(dest == "DTW") |>
  summarize(mean_delay = mean(arr_delay, na.rm = TRUE)) |>
  collect()
```

## Not everything translates

Not every R function has a SQL equivalent. So divide the labor:

* filter / join / group / aggregate **in the database** (shrinks the data)
* `collect()` the small result
* custom functions, modeling, **plotting** -- in R:

```{r}
tbl(con, "flights") |>
  group_by(origin) |>
  summarize(mean_delay = mean(dep_delay, na.rm = TRUE)) |>
  collect() |>                       # <- leave the database here
  ggplot(aes(origin, mean_delay)) +  # <- pure R from here on
  geom_col()
```

## Not all databases are alike

Databases differ along a few axes:

>* **Embedded vs. client-server** -- in-process (SQLite, DuckDB) vs. a server you
>  connect to (Postgres, BigQuery).
>* **Row vs. column** -- row store = transactions (**OLTP**); column store =
>  analytics (**OLAP**).
>* **Relational vs. not** -- most speak SQL; some (document, key-value, graph)
>  don't.

SQLite is an embedded **row** store. For analytics, the fit is an embedded
**column** store -- **DuckDB**.

## A different engine: DuckDB

**DuckDB** -- "SQLite for analytics": embedded (no server), but
**column-oriented** and built for big scans and aggregates. Same `DBI`
interface; only `dbConnect()` changes:

```{r}
duck <- dbConnect(duckdb::duckdb(), ":memory:")   # the one line that changes
dbWriteTable(duck, "mpg", ggplot2::mpg)           # same dbWriteTable() as before
dbListTables(duck)
```

## Same SQL, same `dbplyr`, different engine

The opening query runs on DuckDB verbatim:

```{r}
dbGetQuery(duck, "
  SELECT manufacturer, AVG(hwy) AS mean_hwy
  FROM mpg
  WHERE year = 2008
  GROUP BY manufacturer
  ORDER BY mean_hwy DESC
  LIMIT 5
")
```

And so does `dbplyr`:

```{r}
tbl(duck, "mpg") |>
  group_by(drv) |>
  summarize(mean_hwy = mean(hwy, na.rm = TRUE)) |>
  arrange(desc(mean_hwy)) |>
  collect()
```

Switching engines changed *one line*.

## Query files directly

DuckDB can query a **file on disk directly** -- no `dbWriteTable`, no loading.
Point `FROM` at the file:

```{r}
dbGetQuery(duck, "
  SELECT class, COUNT(*) AS n, ROUND(AVG(hwy), 1) AS mean_hwy
  FROM 'mpg.csv'
  GROUP BY class
  ORDER BY n DESC
  LIMIT 5
")
```

The file *is* the table.

## Exercise: query a file with DuckDB

Open a DuckDB connection and query `mpg.csv` **directly** (no `dbWriteTable`):
the count and average `hwy` for each `drv`, best average first.

```{r ex-duck-file, exercise = TRUE}
duck <- dbConnect(duckdb::duckdb(), ":memory:")
dbGetQuery(duck, "

")
```

```{r ex-duck-file-solution}
duck <- dbConnect(duckdb::duckdb(), ":memory:")
dbGetQuery(duck, "
  SELECT drv, COUNT(*) AS n, ROUND(AVG(hwy), 1) AS mean_hwy
  FROM 'mpg.csv'
  GROUP BY drv
  ORDER BY mean_hwy DESC
")
```

## Parquet: the columnar file format

**Parquet** -- columnar, compressed, typed. A query reads **only the columns it
needs** (CSV forces reading every row, as text). Write it, then query it:

```{r}
# export mpg to a Parquet file on disk
dbExecute(duck, "COPY (SELECT * FROM mpg) TO 'mpg.parquet' (FORMAT PARQUET)")

# query the .parquet file directly
dbGetQuery(duck, "
  SELECT manufacturer, ROUND(AVG(hwy), 1) AS mean_hwy
  FROM 'mpg.parquet'
  WHERE year = 2008
  GROUP BY manufacturer
  ORDER BY mean_hwy DESC
  LIMIT 5
")
```

Modern analytics is often just **a folder of Parquet files** queried by DuckDB.

## A note on Arrow

**Arrow** is the in-memory columnar standard -- tools pass columnar data with no
copy. DuckDB is built on it; Parquet is its on-disk form. Columnar **in memory
(Arrow), on disk (Parquet), in the engine (DuckDB)** is the backbone of modern
analytics -- all driven by the SQL and `dplyr` you already know.

## When data outgrows one machine

Too big for one machine? **Shard** it -- split the table across many machines;
the database routes each query and merges the results. Warehouses (BigQuery,
Snowflake, Redshift) do this for you. From the outside: still **just SQL**.

## Which database when?

| You want to...                                    | Reach for           |
|---------------------------------------------------|---------------------|
| Learn SQL; a small app's local storage            | **SQLite**          |
| Analyze big local tables / Parquet files, fast    | **DuckDB**          |
| A shared database many apps read & write          | Postgres / MySQL    |
| Query an organization's enormous warehouse        | BigQuery, Snowflake |

From R, all are `dbConnect()` + `DBI` + `dbplyr` -- learn the interface once.

## The real payoff: SQL is portable

Look again at what `dbGetQuery()` sent:

```{r}
dbGetQuery(con, "
  SELECT manufacturer, AVG(hwy) AS mean_hwy
  FROM mpg WHERE year = 2008
  GROUP BY manufacturer ORDER BY mean_hwy DESC LIMIT 5
")
```

That string is **not R** -- it's the database's language. It transfers to any
host language that can open a database.

## The same SQL, from Python

`sql_from_python.py` (in this folder, Python standard library only) loads the
same `mpg.csv` and runs the **identical SQL**:

```python
import sqlite3
con = sqlite3.connect(":memory:")
# ... load mpg.csv into a table called `mpg` ...

con.execute("""
    SELECT manufacturer, AVG(hwy) AS mean_hwy
    FROM mpg
    WHERE year = 2008
    GROUP BY manufacturer
    ORDER BY mean_hwy DESC
    LIMIT 5
""").fetchall()
```

The host changed (`con.execute`), the SQL didn't.

## It really runs

`python3 sql_from_python.py` output -- checked in, printed here live:

```{r}
writeLines(readLines("sql_from_python_output.txt"))
```

Same answer as R. **That is the value of SQL**: one query language, any host.

## Summary

* Databases: too big for memory, persistent, shared, and fast (**indexes**).
* `DBI`: `dbConnect()` / `dbWriteTable()` / `dbGetQuery()`; only `dbConnect()` is
  backend-specific.
* **SQL**: one `SELECT` (`FROM`/`WHERE`/`GROUP BY`/`HAVING`/`ORDER BY`/`LIMIT`),
  plus `JOIN`, `WITH`, and window functions -- almost one-to-one with `dplyr`.
* `dbplyr`: `dplyr` → SQL, **lazy**; `collect()` to pull, `show_query()` to see.
* Pattern: do the work in the database, `collect()` the small result.
* Engines vary (row/column, OLTP/OLAP); **DuckDB** queries **CSV/Parquet**
  directly. Parquet (disk) + Arrow (memory) = the columnar backbone.
* SQL is **portable** across host languages.

Reading: [R for Data Science (2e), "Databases"](https://r4ds.hadley.nz/databases),
the [`dbplyr` website](https://dbplyr.tidyverse.org/), and
[DuckDB's R docs](https://duckdb.org/docs/api/r).
