---
title: "lab07 Joins"
author: "Your Name"
date: "2024-03-09"
output: pdf_document
---


```{r setup, echo=FALSE, include=FALSE}
library(tidyverse)
```


# Joins

This section of this lab is about **joins**.

### Overview

In data science, we often find that related information is spread across
multiple data tables. A good example of this is the data from the
`nycflights13` package. As you've seen in lecture, there are actually
five related tables in `nycflights13`: (1) `flights`, (2) `weather`, (3)
`planes`, (4) `airlines`, and (5) `airports`.

We may want to analyze data from two, three, four, or even all five of
these data frames at the same time, as this would allow us to uncover
patterns and relationships that we could not observe in each one
individually. For example, how are flights affected by weather? Do
different airports and airlines react differently to bad weather
conditions?

Fortunately, there are certain variables (i.e., **keys**) that appear in
more than one of the `nycflights13` data tables and uniquely identify
observations in at least one of them. We can use these variables to
**join** the tables together. There are a few different types of joins
--- here's a hierarchy of all of them:

-   Mutating joins
    -   Inner joins
        -   `inner_join()`
    -   Outer joins
        -   `full_join()`
        -   `left_join()`
        -   `right_join()`
-   Filtering joins
    -   Semi-joins
        -   `semi_join()`
    -   Anti-joins
        -   `anti_join()` 


### Load in the data


We're going to take a break from `nycflights13` in this lab. Instead,
we'll work with some of the baseball data frames from the `Lahman`
package: 

[Lahman
documentation](https://cran.r-project.org/web/packages/Lahman/Lahman.pdf)

```{r}
# install.packages("Lahman")
library(Lahman)
```
We'll mostly be working with the `People`, `Batting`, and `Salaries`
data frames. 

### Keys

A key is a variable that connects a pair of data frames and allows us to
join them together. More specifically,

-   A **primary key** is a variable (or set of variables) that uniquely
    identifies each observation.
    -   There is at most one row in the table that corresponds to a
        particular value of the primary key.
    -   Sometimes the observations in a table are uniquely identified by
        more than one variable --- in this case, the set of uniquely
        identifying variables is called a **compound key**.
-   A **foreign key** is a variable (or set of variables) that
    corresponds to a primary key in another table.
    -   In a well-designed database, primary keys and their
        corresponding foreign keys have the same names.

Let's look at some examples.

In the `People` data frame, each row is uniquely identified by
`playerID`, so `playerID` is a **primary key** for `People`. One way to
check this is to count the number of rows in the table for each
`playerID` and then filter to rows where $n > 1$. If we've correctly
identified the primary key, then we will find no `playerID`s for which
$n > 1$:
```{r}
People %>%
  head(5)
```

```{r}
People %>%
    count(playerID) %>%
    filter(n > 1)
```

There's also a `playerID` column in the `Batting` table. We say that
`Batting$playerID` is a **foreign key** for `People$playerID` since each
`playerID` in `Batting` corresponds to one `playerID` in `People`.

However, `playerID` is *not* a primary key for `Batting`. For some
`playerID`s, there are multiple rows in `Batting`:
```{r}
Batting %>%
  head(5)
```


```{r}
Batting %>%
    count(playerID) %>%
    head(5)
```


This makes sense because each row in `Batting` contains the statistics
for one player in one season, and most players' careers last longer than
just one season.


So what is the primary key for `Batting`? After reading through the
Lahman documentation,
we can infer that each row in `Batting` is uniquely identified by the
combination of `playerID`, `yearID`, and `stint`. These three variables
form a **compound key**, and this compound key is a primary key for
`Batting`:

```{r}
Batting %>%
    count(playerID, yearID, stint) %>%
    filter(n > 1)
```

### Mutating joins

Now that we've seen how the tables in the `Lahman` database are
connected by keys, we can practice using joins to actually connect them.

The first category of joins you should know about are **mutating
joins**, which match observations between two data frames and then
$\underline{\text{add variables}}$ from one data frame to the other.
There are four types of mutating joins: (1) inner, (2) left, (3) right,
and (4) full. The latter three are often called outer joins.

#### Mutating join #1: `inner_join()`
Suppose we have two data frames **x** and **y**.

**`inner_join(x, y)`** preserves all rows in **x** that have a matching
key in **y**, and it adds the columns of **y** to **x**.

Let's test this out with the `Batting` and `Salaries` data frames. Note
that `Batting` has 112,184 rows and 22 columns, while `Salaries` has
26,428 rows and 5 columns:

```{r}
glimpse(Batting)
glimpse(Salaries)
```


Note also that there are four overlapping columns between these two data
frames: `yearID`, `teamID`, `lgID`, and `playerID`.

```{r}
bs_inner <- Batting %>%
    inner_join(Salaries)
head(bs_inner)
# Equivalently:
# inner_join(Batting, Salaries)
```

```{r}
dim(bs_inner)
```

The resulting table has 25,441 rows (note that this is less than the
number of rows in each of the original data frames). It has 23 columns
--- the 22 columns in `Batting` plus the one column in `Salaries` that
does not also appear in `Batting`.

Notice that the function automatically joined the two data frames by the
overlapping columns listed above --- this is called a **natural join**.
The four overlapping columns form a compound key.

Natural joins usually give us what we want, but they don't work well if

1.  A column in **x** has the same name as a column in **y** but the two
    columns measure something different or
2.  A column in **x** measures the same thing as a column in **y** but
    the two columns have different names.

In this case, the natural join works well. To manually set the key
(which is not necessary in this example), we can use the `by` argument:

```{r}
bs_inner2 <- Batting %>%
    inner_join(Salaries, by = c("playerID", "yearID", "teamID", "lgID"))
head(bs_inner2)
```

#### Mutating join #2: `left_join()`

**`left_join(x, y)`** preserves all rows in **x**, and it adds the
columns of **y** to **x**.

```{r}
bs_left <- Batting %>%
    left_join(Salaries)
head(bs_left)
```
```{r}
dim(bs_left)
```

The resulting table has 112,184 rows, which is the same number of rows
as `Batting`. This makes sense, as left joins preserve all the rows in
the left-hand data frame.

It also has 23 columns, just like the `inner_join()` table above. This
also makes sense --- we should get the same number of columns for all of
these mutating joins since we are just adding the (unique) columns of
`Salaries` to `Batting`.

Notice that there are a lot of `NA`s in the salary column. This is
because the `Batting` table records data from 1871-2022, but 
`Salaries` only records data from 1985-2016:

```{r}
Batting %>%
  distinct(yearID) %>%
  pull(yearID)
```

```{r}
Salaries %>%
  distinct(yearID) %>%
  pull(yearID)
```

```{r}
Batting %>%
    left_join(Salaries) %>%
    group_by(yearID) %>%
    summarize(prop_NA = mean(is.na(salary))) %>%
    filter(prop_NA < 1) %>%
    pull(yearID) # List of years where at least some salaries were recorded
```


#### Mutating join #3: `right_join()`


**`right_join(x, y)`** preserves all rows in **y**, and it adds the
columns of **y** to **x**.

Here's what a right join looks like for `Batting` and `Salaries`:

```{r}
bs_right <- Batting %>%
    right_join(Salaries)
head(bs_right)
# Equivalently:
# right_join(Batting, Salaries)

# # Also equivalent:
# left_join(Salaries, Batting)
```

```{r}
bs_right2 <- Batting %>%
    right_join(Salaries) %>%
    count(playerID, yearID, teamID, lgID, sort = TRUE)
head(bs_right2)
```

```{r}
dim(bs_right)
```


The resulting data frame has 26,437 rows, which is almost the same
number of rows as `Salaries`. And once again, it has 23 columns.

This time, we see a lot of missing batting statistics toward the bottom
of the table. This is because `Salaries` records data for both batters
and pitchers, but `Batting` only records data for batters.


#### Mutating join #4: `full_join()`

**`full_join(x, y)`** preserves all rows in **x** and **y**, and it adds
the columns of **y** to **x**. 
```{r}
bs_full <- Batting %>%
    full_join(Salaries)
head(bs_full)
# Equivalently:
# full_join(Batting, Salaries)
```

```{r}
dim(bs_full)
```


The resulting table has 23 columns and 113,180 rows, which is more than
the number of rows in either of the original data frames. More
specifically, 113,180 is equal to the following:
```{r}
26437 +       # Number of rows in right_join
  112184 -  # Number of rows in left_join
  25441     # Number of rows in inner_join
```


#### Filtering join #1: `semi_join()`


**`semi_join(x, y)`** preserves all rows in **x** that have a matching
key in **y**. It's kind of like an inner join except it doesn't add any
of the columns from **y**.

For example, here's a semi-join for `Batting` and `Salaries`:

```{r}
bs_semi <- Batting %>%
    semi_join(Salaries)
head(bs_semi)
# Equivalently:
# semi_join(Batting, Salaries)
```

```{r}
dim(bs_semi)
```

This gives us a data frame with the same 22 columns as `Batting`, but
only 25,441 rows (as opposed to the 112,184 rows in `Batting`).



#### Filtering join #2: `anti_join()`


**`anti_join(x, y)`** preserves all rows in **x** that do not have a
matching key in **y**. It does not add any of the columns from **y**.

Let's try an anti-join with `Batting` and `Salaries`:

```{r}
bs_anti <- Batting %>%
    anti_join(Salaries)
head(bs_anti)
# Equivalently:
# anti_join(Batting, Salaries)
```

```{r}
dim(bs_anti)
```

Similar to the semi-join from above, we get a data frame with the same
22 columns as `Batting`. However, the anti-join table has 86,743 rows.
This means that 86,743 rows in `Batting` have no matching key in
`Salaries`, and the remaining 25,441 rows in `Batting` have at least one
matching key in `Salaries`.

### Exercise 1

**`Batting` contains single-season batting statistics for individual
batters. `Salaries` contains single-season salary information for
individual players. `People` contains demographic information about
individual players, including their name and where they are from.**

**Using the information in these three data frames, create a scatterplot
where the x-axis is `salary` and the y-axis is `HR` (home runs). Include
only data from the 1998 season for batters who played at least 100 games
(`G >= 100`) and who were born in either the United States (USA) or the
Dominican Republic (D.R.).**

**BONUS: There are only four players in this scatterplot who hit at
least 40 home runs and made at least \$7,500,000 in 1998. Can you find
their first and last names?**

```{r}

```



**SOLUTION**
```{r}

table <- Batting %>%
            inner_join(Salaries) %>%
            inner_join(People) %>%
            filter(yearID == 1998, birthCountry %in% c("USA", "D.R."), G >= 100)

table %>%
    ggplot() +
    geom_point(aes(x = salary, y = HR),
                   col = "darkslategray", fill = "cornflowerblue",
                   size = 3, stroke = 1.5, shape = "circle filled") +
    theme_bw() +
    ggtitle("Home runs vs. salary for US and DR players in 1998") +
    xlab("Salary") + ylab("Home runs")

table %>% filter(salary > 7500000, HR > 40) %>% select(nameFirst, nameLast, HR, salary)
```



### Exercise 2

**At the end of every baseball season, Gold Glove awards are handed out
to the best defensive players at each position. This information, along
with other awards data, is stored in the `AwardsPlayers` data frame. The
following code creates a tibble of the `playerID`s of players who have
won a Gold Glove since 1950:**

```{r}
gold_gloves <- AwardsPlayers %>%
                filter(awardID == "Gold Glove", yearID >= 1950) %>%
                select(playerID)
```

**The `People` data frame contains demographic information about
individual players, including their birth months. Filter `People` to
only players who have won a Gold Glove award. What is the most common
birth month of these Gold Glove winners? The least common?**

**BONUS: Filter `People` to only players who have *not* won a Gold Glove
award. What is the most common birth month of these non-Gold Glove
winners? The least common?**


```{r}

```


**SOLUTION**
```{r}
#@title Solution to exercise 2

People %>%
    semi_join(gold_gloves) %>%
    filter(!is.na(birthMonth)) %>%
    group_by(birthMonth) %>%
    summarize(n = n()) %>%
    arrange(desc(n))
# Most common is April, least common is February

People %>%
    anti_join(gold_gloves) %>%
    filter(!is.na(birthMonth)) %>%
    group_by(birthMonth) %>%
    summarize(n = n()) %>%
    arrange(desc(n))
# Most common is August, least common is June
```
